---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
For the scope of the assignment please see the README.md file

## Loading and preprocessing the data
The data set for this assessment is suplied in the GitHub repository so I will assume that it is already downloaded and present in the directory where the R markdown script is run. However if that is not the case and the data needs to be downloaded manually, then it can be found here: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

  1. Show any code that is needed to

  2. Load the data (i.e. read.csv())

  First unpack the compressed file, then load into a dataframe:
```{r, ReadingData,echo=TRUE}
unzip("./activity.zip")
activityLog<-read.csv("activity.csv")
```

  3. Process/transform the data (if necessary) into a format suitable for your analysis. *Answer: This does not to be necessary and removing NA's or recunstructing data will be done below.*


## What is mean total number of steps taken per day?
 
  1.  Calculate the total number of steps taken per day
  
      First we need to remove all NA's from the dataframe

```{r, CleanAndAggregate, echo=TRUE}
actLogCompCases<-activityLog[complete.cases(activityLog),]
stepsPerDay<-aggregate(actLogCompCases$steps,by=list(actLogCompCases$date),FUN=sum)
```

  2.  If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

```{r, HistogramPlot ,echo=TRUE}
hist(stepsPerDay$x,xlab="number of steps taken per day",ylab="number of days",main="")
```

  3.  Calculate and report the mean and median of the total number of steps taken per day
  
Mean of steps taken per day:
```{r,echo=TRUE}
mean(stepsPerDay$x)
```

Median of steps taken per day:
```{r,echo=TRUE}
median(stepsPerDay$x)
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
